% Aufschreibesysteme Flow Matching — rolling preprint draft.
% Rolling means: this paper is NOT pegged to a conference deadline. PRs that
% only exist to chase a deadline window are out of scope.

\documentclass[10pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\title{Aufschreibesysteme Flow Matching: \\
       Treating Lossy Codecs as First-Class Generative Regimes}
\author{aufschreibesysteme-fm contributors}
\date{rolling preprint, last compiled \today}

\begin{document}
\maketitle

\begin{abstract}
We treat lossy codecs (JPEG, gramophone, film grain, CRT phosphor, \dots)
as first-class generative regimes rather than post-hoc style filters,
attaching to each a measurement-derived \emph{frozen} physical noise
prior and a per-regime LoRA expert. A top-$k$ soft mixture-of-experts
router gates the experts under a Switch-style load-balance loss with a
runtime collapse monitor. On seven regimes spanning the three
\emph{Aufschreibesysteme} epochs (1800 / 1900 / 2000) we measure (M1)
regime-classification recall, (M2) per-regime FID, (M3) cross-regime
morph trajectory smoothness, and (M4) prompt-conditioned inscription
consistency. The architecture, code, weights, and HF Space demo are
released under Apache-2.0.
\end{abstract}

\section{Introduction}\label{sec:intro}

Flow Matching and DiT-style generators implicitly assume a flat,
homogeneous latent space. Real inscription channels are not flat — a
JPEG quantisation lattice, the PSD of a 78\,rpm shellac groove,
ISO12233 film grain, and CRT phosphor decay each impose distinct
measurable structure on the signal. We make those structures
\emph{first-class regimes}: each one carries its own measurement-derived
prior, its own LoRA expert, and shares a routing gate. The framing is
Kittlerian (\S\ref{sec:kittler}) but the metrics are signal-materiality
only (\S\ref{sec:experiments}).

\paragraph{Contributions.}
\begin{enumerate}
    \item A frozen, measurement-derived noise-prior architecture that
          treats lossy codecs as regimes, not styles.
    \item A top-2 soft Mixture-of-Experts LoRA router with a runtime
          collapse monitor and a Switch-style load-balance loss.
    \item Cross-regime morphing via Diffusion Schr\"odinger Bridge
          Matching at inference time, without adding to training cost.
    \item An Apache-2.0 release with weights, HF Space, and a 1-command
          reproduction (\verb|afm demo --regime jpeg|).
\end{enumerate}

\section{Related Work}\label{sec:related}
Lossy-codec-aware generation: \cite{ddcm2025,fdm2024}. Channel-coding
diffusion: \cite{diffjscc2024}. Grain-specific generators:
\cite{fganrn2025}. MoE LoRA: \cite{loramixer2025}. Diffusion forcing
for temporal generation: \cite{diffusionforcing2024}. None treat the
codec itself as a first-class regime nor use a frozen physical prior.

\section{Method}\label{sec:method}
\subsection{Regimes}
We use seven regimes spanning the three Aufschreibesysteme epochs:
\emph{parchment}, \emph{typewriter} (1800); \emph{gramophone},
\emph{photograph}, \emph{film} (1900); \emph{jpeg}, \emph{crt} (2000).
Each regime carries a codec, a measurement-derived physical noise prior,
and a $\sigma_{\min}$ floor.

\subsection{Frozen physical prior}
Per-regime PSDs are stored via PyTorch \texttt{register\_buffer}, never
as \texttt{nn.Parameter}, and never have \texttt{requires\_grad} flipped.
Sampling is via the spectral form
\[
\epsilon_r \;=\; \mathcal{F}^{-1}\!\bigl(\sqrt{\mathrm{PSD}_r}\cdot
                                       \mathcal{F}(z)\bigr),
\qquad z \sim \mathcal{N}(0, I).
\]
We deliberately keep this layer non-learnable; the Lacanian caveat is in
the appendix.

\subsection{LoRA experts and routing}
Each regime has a rank-$32$ LoRA adapter attached to the cross-attention
$Q,K,V$ projections and the FFN of the base DiT. A 2-layer MLP gate
(hidden 256) consumes
$(\text{layer-4 patch-mean}\oplus\text{text CLS})$ and emits a softmax
distribution over the seven regimes; we keep the top-$2$ weights and
re-normalise. A Switch-style load-balance loss
$\mathcal{L}_{lb} = \lambda N \sum_r f_r P_r$ with $\lambda = 0.01$ is
augmented by a runtime collapse monitor: if the rolling-window mean
entropy of the gate distribution falls below $0.3$ nats for $1000$
optimiser steps, $\lambda$ doubles.

\subsection{Per-pixel $\sigma$ with prior floor}
We adopt a MuLAN-style per-pixel $\sigma$ head clipped from below:
\[
\sigma_{\mathrm{pred}}(x, r) \;=\; \max\!\bigl(\sigma_{\min}^{(r)},
                                              \mathrm{MuLAN}_\theta(x)\bigr).
\]
This guarantees the noise model respects the regime's physical floor
even when MuLAN's free head would prefer a smaller $\sigma$.

\subsection{Schedulers}
We use rectified Flow Matching (\cite{rfm2022}) for all spatial regimes,
and Diffusion Forcing (\cite{diffusionforcing2024}) per-frame for the
single temporal regime (film).

\subsection{Cross-regime morphing}
\verb|afm morph --from A --to B| runs Diffusion Schr\"odinger Bridge
Matching at inference time. This does \emph{not} add to training cost,
so the YouTube short and HF Space teaser do not change the wall-clock
budget reported below.

\section{Experiments}\label{sec:experiments}
\subsection{Setup}
Base model: SANA-1.6B \cite{sana2024} (Apache-2.0), frozen. Training:
bf16, AdamW8bit, batch $2$ with grad-accum $8$, $30\,000$ steps per
regime, \texttt{save\_steps=500}. Sequential per-regime LoRA fits in
$\approx 18$--$20$\,GB VRAM on RTX 4090. Single-GPU wall-clock for the
seven regimes plus gate + eval + space deploy is approximately $66$
hours.

\subsection{Metrics and baselines}
\textbf{M1}: ResNet-50 fine-tuned for $7$-class regime classification,
held-out split; we report macro recall. \textbf{M2}: per-regime FID
against $5{,}000$ real references each. \textbf{M3}: LPIPS-trajectory
monotonicity and Perceptual Path Length across the $50$-step DSBM morph.
\textbf{M4}: variance of CLIPScore across regimes for the same prompt.
\textbf{Baselines}: B1 vanilla SANA + style-LoRA stack, B2 SANA +
IP-Adapter, B3 DDCM, B4 FGA-NN. All share captions
(\texttt{data/manifests/captions\_eval.txt}), eval seed $17$, step
budget $200\,000$, and the frozen text encoder.

\subsection{Results}
\emph{[Numbers go here once the MVP run completes. The rolling preprint
will be updated; the figure / table layout is fixed.]}

\section{Kittler Motivation (Appendix Pointer)}\label{sec:kittler}
The seven regimes were chosen to span Kittler's three
\emph{Aufschreibesysteme} epochs, with the 1900 triad
(gramophone + photograph + film) intact. The codebase relies on this
framing only for the regime list and the architectural decision to
freeze the physical prior; metrics are signal-materiality only. See
\verb|docs/appendix_kittler.md| in the repo for the full reading,
including the Lacan caveat and references to Winthrop-Young, Peters,
and Kr\"amer.

\section{Limitations}
The MVP is sequential per-regime training; parallel training is
infeasible at 24\,GB VRAM. The DSBM morph uses a single IPF iteration;
the full DSBM lives on the post-MVP roadmap. Audio is restricted to
public-domain whitelist items because EU neighbouring rights run
$70$ years past the performer's death and disqualify much of the
Internet Archive 78\,rpm bulk.

\section{Reproducibility}
Single command:
\begin{verbatim}
afm oss --resume --pause-on=licence-fail,collapse,space-deploy,preprint \
        --config configs/mvp.yaml
\end{verbatim}
Four human-review gates (license audit, expert collapse, HF Space
publish, preprint draft) pause for sign-off and resume cleanly from
\texttt{state.json}. Apache-2.0 throughout.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
